---
title: "Insurance and no insurance"
author: "Erik Ø. Sørensen"
date: "August 4, 2015"
output: github_document
---

This document is to estimate on the insurance-no-insurance data. The editor wants us to comment on
whether sigma is the same or different between those who bought and those who didn't buy
insurance. The document was changed in a minor way July 20th 2017 to 1) remove internal discussions
about response to the editor in the revision process and 2) addition of version information at the end. Minor change October 13, 2018,
in order to render markdown output in GitHub format.

# Setup - loading library and subsetting data
First I need to extract and run the library
```{r, echo=FALSE, warning=FALSE, results='hide'}
source("luck-model.R")
```

Data subsetting (the global datasets are defined in `luck-model`):
```{r}
ds.noin <- datastruct(players[players$round==1 & players$insurance==0,], situations)
ds.in   <- datastruct(players[players$round==1 & players$insurance==1,], situations)
```

For the likelihood ratio test we also need the full dataset of round 1.
```{r}
ds <- datastruct(players[players$round==1,], situations)
```



# On the insured
Defining the likelihood function on this subset of data:

```{r}
ll.in <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8) {
    pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma)
    ll.luck(ds.in, pvector)
}
```

Without constraints: 
```{r }
fit.in1 <- mle(minuslog = ll.in, fixed=list(b1.c=0, b5.c=-1000), 
              method="BFGS")
summary(fit.in1)
knitr::kable(delta.method(grtrans, vcov(fit.in1), coef(fit.in1)))
logLik(fit.in1)
```

Now constraining the LE share to be zero:
```{r }
fit.in2 <- mle(minuslog = ll.in, fixed=list(b1.c=0,b3.c=-1000, b5.c=-1000), 
              method="BFGS")
summary(fit.in2)
knitr::kable(delta.method(grtrans, vcov(fit.in2), coef(fit.in2)))
logLik(fit.in2)
```



# On the non-insured

Defining the likelihood function on this subset of data:
```{r}
ll.noin <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8) {
    pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma)
    ll.luck(ds.noin, pvector)
}
```

First the model in without constraints:
```{r }
fit.noin1 <- mle(minuslog = ll.noin, 
               fixed=list(b1.c=0, b5.c=-1000), method="BFGS")
summary(fit.noin1)
knitr::kable(delta.method(grtrans, vcov(fit.noin1), coef(fit.noin1)))
logLik(fit.noin1)
```

Now constraining the LE share to be zero:
```{r }
fit.noin2 <- mle(minuslog = ll.noin, 
               fixed=list(b1.c=0, b3.c=-1000, b5.c=-1000), method="BFGS")
summary(fit.noin2)
knitr::kable(delta.method(grtrans, vcov(fit.noin2), coef(fit.noin2)))
logLik(fit.noin2)
```


# Constraining $\sigma$ to be the same

In order to use the variant likelihood functions, we need a function
that indicates which group we are are in:
```{r}
setGeneric("insured", function(s) standardGeneric("insured"))
setMethod("insured", "individual",
          function(s) {
            if (s@insurance==1)
              1
            else
              2
          }
)
```

Defining the likelihood function on all data and with the group indicator function:

```{r}
ll.in.2K <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8,
                     b1.d=0.01, b2.d=0.01, b3.d=0.01, b4.d=0.01, b5.d=0.01, mu.d=0.01, lsigma.d=0.01) {
    pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma, b1.d, b2.d, b3.d, b4.d, b5.d, mu.d, lsigma.d)
    ll.luck.2K(ds, pvector, insured)
}
```

Now estimating on the unconstrained model to confirm that the variant estimation command works:
```{r }
fit.2K <- mle(minuslog = ll.in.2K, 
              start=list(b2.c=-0.57, b4.c=-0.05, mu=-0.93, lsigma=0.68,
                         b2.d=2.3, b4.d=0.58, mu.d=0.06, lsigma.d=1.8),
              fixed=list(b1.c=0, b1.d=0, b3.c=-1000, b3.d=0, b5.c=-1000, b5.d=0),
              method="BFGS")
summary(fit.2K)
logLik(fit.2K)
```

We see that these numbers conform to those estimated seperately above, so this seems like a success.

## Estimating with constrained $\sigma$

First with non-constrained share of luck egalitarians:
```{r }
fit.2K.1 <- mle(minuslog = ll.in.2K, 
              start=list(b2.c=-0.57, b3.c=-0.9, b4.c=-0.05, mu=-0.93, lsigma=1.2,
                         b2.d=2.3, b3.d=0.1, b4.d=0.58, mu.d=0.06),
              fixed=list(b1.c=0, b1.d=0, b5.c=-1000, b5.d=0, lsigma.d=0),
              method="BFGS")
summary(fit.2K.1)
logLik(fit.2K.1)
```

Now with the share of luck egalitarians constrained to zero:
```{r}
fit.2K.2 <- mle(minuslog = ll.in.2K, 
              start=list(b2.c=-0.57, b4.c=-0.05, mu=-0.93, lsigma=1.2,
                         b2.d=2.3, b4.d=0.58, mu.d=0.06),
              fixed=list(b1.c=0, b1.d=0, b3.c=-1000, b3.d=0, 
                         b5.c=-1000, b5.d=0, lsigma.d=0),
              method="BFGS")
summary(fit.2K.2)
logLik(fit.2K.2)
```

## Tests and pvalues of common $\sigma$
First testing column 1 and 3 are the same.
```{r}
diff1 <- -2*(logLik(fit.2K.1) - (logLik(fit.in1)  + logLik(fit.noin1)))
diff1 
( p_non_constrained <- pchisq(diff1, df=1, lower.tail=FALSE))
```

Now testing with the share of LE constrained to zero (2 and 4):
```{r}
diff2 <- -2*(logLik(fit.2K.2) - (logLik(fit.in2)  + logLik(fit.noin2)))
diff2
( p_constrained <- pchisq(diff2, df=1, lower.tail=FALSE))
```

Standard analytic inference on maximum likelihood estimates is based on 
asymptotic theory, and asymptotically, confidence
intervals in the interior of the parameter space are symmetric. But using only the 
local information at the unconstrained estimates might be restrictive, and we now
include a footnote with likelihood ratio test for the equality of $\sigma$ between 
these specifications (which cannot be rejected). Since the p-values are far from being
close to significant, we believe that developing small-sample properties
of $\sigma$ will not contribute much to the paper.

Footnote for paper: The constraint that the $\sigma$s are the same in column 1 and 3
and 2 and 4 cannot be rejected (likelihood ratio tests based on a pooled
and restricted model: $\chi_1^2=0.56, p=0.45$ and $\chi_1^2=1.02, p=0.31$, 
respectively).

# Session Info
This documents the versions used of R and the involved packages to calculate these numbers:

```{r}
sessionInfo()
```