---
title: "Main table"
author: "Erik Ø. Sørensen"
date: "07. feb. 2015"
output: 
  html_document:
    theme: readable
---

This document is to estimate the specifications in the main table. Make
sure to first extract the R-code from `luck-model.Rmd` (using `knitr::purl('luck-model.Rmd')`,
unfortunately, knitr cannot call itself to do it automatically). Minor change July 20th, 2017 
to document the version of packages used at the end of the document.

# Setup - loading library and data
First I need to extract and run the library
```{r, echo=FALSE, warning=FALSE, results='hide'}
source("luck-model.R")
```
For estimation, first look at the first round of data.
```{r }
ds.luck1 <- datastruct(players[players$round==1,], situations)
```

Next, define the likelihood function.
```{r} 
ll <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8) {
     pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma)
    ll.luck(ds.luck1, pvector)
}
```


# Specifications
First, estimate the main specification. 
```{r}
fit.all <- mle(minuslog = ll, fixed=list(b1.c=0, b5.c=-1000), 
               method="BFGS", control=list(trace=TRUE, REPORT=2))
summary(fit.all)
logLik(fit.all)
knitr::kable(delta.method(grtrans, vcov(fit.all), coef(fit.all)))

```

The ideals are, in order: SE, L, LE, CC (and sCC, not estimated in this document).

Remove LE:
```{r}
fit.2 <- mle(minuslog = ll, fixed=list(b1.c=0, b3.c=-1000, b5.c=-1000), method="BFGS",
             control=list(trace=TRUE, REPORT=2))
summary(fit.2)
knitr::kable(delta.method(grtrans, vcov(fit.2), coef(fit.2)))
logLik(fit.2)
```

Now, remove LE and SE. When I remove one of the substantial ideals, the likelihood becomes
quite ill-behaved and sensitive to starting values. These starting values converge at
the best (local) point of convergence that I have found.
```{r}
fit.3 <- mle(minuslog = ll, start=list(b4.c=-0.292, mu=-10.8, lsigma=2.398), 
             fixed=list(b1.c=-1000, b2.c=0, b3.c=-1000, b5.c=-1000), method="BFGS",
             control=list(trace=TRUE, REPORT=2))
summary(fit.3)
knitr::kable(delta.method(grtrans, vcov(fit.3), coef(fit.3)))
logLik(fit.3)
```

Remove LE and L:
```{r}
fit.4 <- mle(minuslog = ll, start=list(b4.c=0.48, mu=-1.3, lsigma=1.9), 
             fixed=list(b1.c=0, b2.c=-1000, b3.c=-1000, b5.c=-1000),
             method="BFGS", control=list(trace=TRUE, REPORT=2))
summary(fit.4)
knitr::kable(delta.method(grtrans, vcov(fit.4), coef(fit.4)))
logLik(fit.4)
```

Remove LE and CC:
```{r}
fit.5 <- mle(minuslog = ll, start=list(b2.c=0.034, mu=-1.6, lsigma=1.0),
             fixed=list(b1.c=0, b3.c=-1000, b4.c=-1000, b5.c=-1000), method="BFGS",
             control=list(trace=TRUE, REPORT=2))
summary(fit.5)
bhhh.fit5 <- vcov.bhhh(ds.luck1, fit.5)
knitr::kable(delta.method(grtrans, vcov(fit.5), coef(fit.5)))
bhhh.fit5 <- vcov.bhhh(ds.luck1, fit.5)
logLik(fit.5)
```

# Different way of calculating standard errors
The above estimates of standard errors are based on the defaults of the stats4 library,
which we cite in the paper. Some standard errors (for $\sigma$ are not well estimated). 
One could think that this results from how the stats4 library calculates 
standard errors: Based, as far as I understand, on taking the direct numerical Hessian
of the log likelihood function. An alternative is the BHHH outer product of the gradient.
I do that here (and conclude from this that it doesn't really matter which way we do it). 

```{r}
bhhh.all <- vcov.bhhh(ds.luck1, fit.all)
knitr::kable(delta.method(grtrans, bhhh.all, coef(fit.all)))
bhhh.fit2 <- vcov.bhhh(ds.luck1, fit.2)
knitr::kable(delta.method(grtrans, bhhh.fit2, coef(fit.2)))
bhhh.fit3 <- vcov.bhhh(ds.luck1, fit.3)
knitr::kable(delta.method(grtrans, bhhh.fit3, coef(fit.3)))
bhhh.fit4 <- vcov.bhhh(ds.luck1, fit.4)
knitr::kable(delta.method(grtrans, bhhh.fit4, coef(fit.4)))
bhhh.fit5 <- vcov.bhhh(ds.luck1, fit.5)
knitr::kable(delta.method(grtrans, bhhh.fit5, coef(fit.5)))
```

# Session Info
This documents the versions used of R and the involved packages to calculate these numbers:
```{r}
sessionInfo()
```



