---
title: "Pooling with extension"
author: "Erik Ø. Sørensen"
date: "August 7, 2015"
output: github_document
---

We have a table in the paper in which we pool the preferred data with the extension. This file generates the estimates
and tests for equality of parameters across these experiments. 
The document was changed in a minor way July 20th 2017 to 1) remove internal discussions
about response to the editor and referees in the revision process and 
2) addition of version information at the end.
Minor change October 13, 2018,
in order to render markdown output in GitHub format.


First I need to extract and run the library, load the data
```{r, echo=FALSE, warning=FALSE, results='hide'}
source("luck-model.R")
ds1 <- datastruct(players[players$round==1,], situations)
ds2 <- datastruct(players[players$round==2,], situations)
ds.all <- datastruct(players, situations)
```

# Estimation on the extension data
```{r}
ll.2 <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8) {
    pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma)
    ll.luck(ds2, pvector)
}
```
Without constraints: 
```{r }
fit.later1 <- mle(minuslog = ll.2, fixed=list(b1.c=0, b5.c=-1000), 
              method="BFGS")
summary(fit.later1)
knitr::kable(delta.method(grtrans, vcov(fit.later1), coef(fit.later1)))
logLik(fit.later1)
```

Now constraining the LE share to be zero:
```{r }
fit.later2 <- mle(minuslog = ll.2, fixed=list(b1.c=0,b3.c=-1000, b5.c=-1000), 
              method="BFGS")
summary(fit.later2)
knitr::kable(delta.method(grtrans, vcov(fit.later2), coef(fit.later2)))
logLik(fit.later2)
```

# Estimation on the pooled data.
```{r}
ll.all <- function(b1.c=0, b2.c=-0.06, b3.c=-3, b4.c=0.1, b5.c=-1000, mu=-0.85, lsigma=0.8) {
    pvector <- c(b1.c, b2.c, b3.c, b4.c, b5.c, mu, lsigma)
    ll.luck(ds.all, pvector)
}
```

Without constraints: 
```{r }
fit.all1 <- mle(minuslog = ll.all, fixed=list(b1.c=0, b5.c=-1000), 
              method="BFGS")
summary(fit.all1)
knitr::kable(delta.method(grtrans, vcov(fit.all1), coef(fit.all1)))
logLik(fit.all1)
```

Now constraining the LE share to be zero:
```{r }
fit.all2 <- mle(minuslog = ll.all, fixed=list(b1.c=0,b3.c=-1000, b5.c=-1000), 
              method="BFGS")
summary(fit.all2)
knitr::kable(delta.method(grtrans, vcov(fit.all2), coef(fit.all2)))
logLik(fit.all2)
```

# Testing pooling

First testing for the unconstrained model:
```{r}
diff1 <- -2*( logLik(fit.all1) - (-817.0486 + logLik(fit.later1)))
diff1
pchisq(diff1, df=6, lower.tail=FALSE)
```

Now testing with the share of LE constrained to zero:
```{r}
diff2 <- -2*(logLik(fit.all2)  - (-817.836 + logLik(fit.later2)))
diff2
pchisq(diff2, df=5, lower.tail=FALSE)
```

For footnote to paper: We cannot reject the restricted models in columns (3) and (4) against
the unrestricted one in column (1) and (2) of table 9 and (1) and (2) of table 6 ($\chi^2_6=5.67, p=0.46$ for
the model with LE included, $\chi^2_5=5.44, p=0.36$ with the lambda^LE=0).


# Session Info
This documents the versions used of R and the involved packages to calculate these numbers:
```{r}
sessionInfo()
```
